---
title: "ANOVA Analysis of Tooth Growth Data"
author: "Bharani"
date: "`r Sys.Date()`"
format: html
execute:
  echo: true  
  output: true  
  warning: false  
  message: false  
---

# Introduction

This report presents an **ANOVA analysis** of the `ToothGrowth` dataset in R.
We consider the **ToothGrowth** dataset from the **biomedical research domain**. It explores how different **dosages of Vitamin C** and **supplement types** affect **tooth growth** in guinea pigs. We examine the effect of **dose levels** and **supplement type** on tooth length. The analysis includes:

-   **Exploratory Data Analysis (EDA)**
-   **Assumption Testing (Normality & Homogeneity of Variance)**
-   **One-Factor and Two-Factor ANOVA**
-   **Post-hoc Comparisons (Tukey's HSD)**
-   **Conclusions & Interpretations**

------------------------------------------------------------------------

### **Exploratory Data Analysis (EDA)**

\# \*\*Introduction\*\*

The \`ToothGrowth\` dataset in R contains \*\*60 observations\*\* measuring the \*\*effect of Vitamin C on tooth growth in guinea pigs\*\*. The experiment was conducted using \*\*two supplement types\*\* at \*\*three different dose levels\*\*.

##### **Description of Variables**

| Variable | Description |
|---------------------------------|---------------------------------------|
| `len` | Tooth length (numeric) |
| `supp` | Supplement type (`VC` for Vitamin C, `OJ` for Orange Juice) |
| `dose` | Dose level of supplement (numeric, but often treated as categorical: `0.5`, `1`, `2` mg) |

```{r}
# Data Loading and Preparation

# Load necessary libraries
library(ggplot2)
library(car)
library(dplyr)

# Load the dataset
data("ToothGrowth")

# Create a local copy
tooth_data <- ToothGrowth

# Convert dose and supp to factors
tooth_data$dose <- as.factor(tooth_data$dose)
tooth_data$supp <- as.factor(tooth_data$supp)

# View summary statistics
summary(tooth_data)

```

Boxplots help us visualize the distribution of **tooth length (len)** across different **dose levels** and **supplement types (supp)**. Below, we will explain what these boxplots tell us about the data.

```{r}
# Boxplot of Tooth Length by Dose
ggplot(tooth_data, aes(x = dose, y = len, fill = dose)) +
  geom_boxplot() +
  labs(title = "Effect of Dose on Tooth Growth", x = "Dose", y = "Tooth Length") +
  theme_minimal()

# Boxplot of Tooth Length by Supplement Type
ggplot(tooth_data, aes(x = supp, y = len, fill = supp)) +
  geom_boxplot() +
  labs(title = "Effect of Supplement Type on Tooth Growth", x = "Supplement Type", y = "Tooth Length") +
  theme_minimal()
```

-   **Each boxplot represents a dose group (0.5 mg, 1 mg, 2 mg).**
-   The **median (middle line in the box)** increases as the dose increases.
-   The **spread (interquartile range, IQR)** of tooth length measurements is larger at **higher doses**.
-   The presence of **outliers (dots outside the whiskers)** suggests some guinea pigs responded more strongly than others.
-   **Conclusion:** **Higher doses generally lead to greater tooth growth.**

### **Assumption Testing (Normality & Homogeneity of Variance)**

Before performing **ANOVA (Analysis of Variance)** on the ToothGrowth dataset, we need to check two key assumptions:

1.  **Normality** – The data should be approximately normally distributed within each group.

2.  **Homogeneity of Variance (Homoscedasticity)** – The variance should be similar across all groups.

**1. Testing for Normality**

**Why?**

• ANOVA assumes that the dependent variable (len = tooth length) is **normally distributed within each group**.

• If this assumption is violated, the **F-test may not be valid**.

• However, ANOVA is **robust** to mild deviations from normality when sample sizes are large.

```{r}
# Perform Shapiro-Wilk test for normality
shapiro_result <- shapiro.test(tooth_data$len)

# Print formatted results
cat("\n------------------------------------------\n")
cat("         SHAPIRO-WILK TEST FOR NORMALITY        \n")
cat("------------------------------------------\n")
cat(sprintf("Test Statistic (W)   : %.5f", shapiro_result$statistic))
cat(sprintf("\nP-Value              : %.5f", shapiro_result$p.value))

# Interpretation based on p-value
if (shapiro_result$p.value > 0.05) {
  cat("\n\nConclusion: The p-value is greater than 0.05, so we fail to reject the null hypothesis.\n")
  cat("This suggests that the data follows a normal distribution. ✅\n")
} else {
  cat("\n\nConclusion: The p-value is less than 0.05, so we reject the null hypothesis.\n")
  cat("This suggests that the data significantly deviates from a normal distribution. ❌\n")
}
cat("\n------------------------------------------\n")

# Generate QQ Plot
qq_plot <- ggplot(tooth_data, aes(sample = len)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "QQ Plot for Normality",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()

# Print the QQ Plot
print(qq_plot)

```

• **Null Hypothesis (H₀):** The data is **normally distributed**.

• **Alternative Hypothesis (H₁):** The data is **not normally distributed**.

• If p \> 0.05, **fail to reject H₀** → Data is normal (assumption holds).

• If p \< 0.05, **reject H₀** → Data is not normal (assumption violated).

**2. Testing for Homogeneity of Variance**

**Why?**

• ANOVA assumes **equal variances** across all groups.

• If variances differ significantly, **F-test results may be biased**.

**Method: Levene’s Test**

```{r}
# Load necessary library
library(car)

# Perform Levene's Test for Dose
levene_dose <- leveneTest(len ~ dose, data = tooth_data)

# Perform Levene's Test for Supplement
levene_supp <- leveneTest(len ~ supp, data = tooth_data)

# Print formatted results
cat("\n------------------------------------------\n")
cat("     LEVENE'S TEST FOR HOMOGENEITY OF VARIANCE    \n")
cat("------------------------------------------\n")

# Print results for Dose
cat("\n📌 Testing Variance Across Dose Groups:\n")
cat(sprintf("Test Statistic (F)   : %.5f", levene_dose$"F value"[1]))
cat(sprintf("\nP-Value              : %.5f", levene_dose$"Pr(>F)"[1]))

# Interpretation for Dose
if (levene_dose$"Pr(>F)"[1] > 0.05) {
  cat("\n\nConclusion: The p-value is greater than 0.05, so we fail to reject the null hypothesis.\n")
  cat("This suggests that the variance is equal across different dose groups. ✅\n")
} else {
  cat("\n\nConclusion: The p-value is less than 0.05, so we reject the null hypothesis.\n")
  cat("This suggests that the variance differs significantly across dose groups. ❌\n")
}

# Print results for Supplement
cat("\n\n📌 Testing Variance Across Supplement Types:\n")
cat(sprintf("Test Statistic (F)   : %.5f", levene_supp$"F value"[1]))
cat(sprintf("\nP-Value              : %.5f", levene_supp$"Pr(>F)"[1]))

# Interpretation for Supplement
if (levene_supp$"Pr(>F)"[1] > 0.05) {
  cat("\n\nConclusion: The p-value is greater than 0.05, so we fail to reject the null hypothesis.\n")
  cat("This suggests that the variance is equal across supplement types (OJ & VC). ✅\n")
} else {
  cat("\n\nConclusion: The p-value is less than 0.05, so we reject the null hypothesis.\n")
  cat("This suggests that the variance differs significantly across supplement types. ❌\n")
}
cat("\n------------------------------------------\n")
```

-   **Null Hypothesis (H₀):** All groups have **equal variance**.

-   **Alternative Hypothesis (H₁):** At least one group has **different variance**.

-   If p \> 0.05, **fail to reject H₀** → Variance is equal (assumption holds).

-   If p \< 0.05, **reject H₀** → Variance is unequal (assumption violated).

-   Since p = 0.385 (\> 0.05), we **do not reject** the null hypothesis.

-   **Conclusion:** Variance is **equal** across dose groups.

### **One-Factor and Two-Factor ANOVA**

**One-Factor ANOVA: Effect of Dose on Tooth Growth**

```{r}
# Perform One-Factor ANOVA for Dose
anova_dose <- aov(len ~ dose, data = tooth_data)
anova_dose_summary <- summary(anova_dose)

# Print formatted output
cat("\n------------------------------------------\n")
cat("     ONE-FACTOR ANOVA: EFFECT OF DOSE     \n")
cat("------------------------------------------\n")
print(anova_dose_summary)

```

**Interpretation:**

**p-value (Pr(\>F)) \< 0.001** → **Dose has a strong, significant effect on tooth growth**. **Higher doses lead to significantly greater tooth growth**.

**One-Factor ANOVA: Effect of Supplement Type**

```{r}
# Perform One-Factor ANOVA for Supplement Type
anova_supp <- aov(len ~ supp, data = tooth_data)
anova_supp_summary <- summary(anova_supp)

# Print formatted output
cat("\n-----------------------------------------\n")
cat("  ONE-FACTOR ANOVA: EFFECT OF SUPPLEMENT  \n")
cat("------------------------------------------\n")
print(anova_supp_summary)
```

**Interpretation:**

**p-value = 0.0604** → **Not significant at the 5% level**. Supplement type **alone** does not significantly impact tooth growth.

**Two-Factor ANOVA: Additive Model (Dose + Supplement)**

```{r}
# Perform Two-Factor ANOVA (without interaction)
anova_additive <- aov(len ~ dose + supp, data = tooth_data)
anova_additive_summary <- summary(anova_additive)

# Print formatted output
cat("\n--------------------------------------------------\n")
cat("   TWO-FACTOR ANOVA: ADDITIVE MODEL (DOSE + SUPP)   \n")
cat("----------------------------------------------------\n")
print(anova_additive_summary)
```

**Interpretation:**

**Both dose (p \< 0.001) and supplement type (p = 0.000429) are significant. Unlike in the one-factor model**, supplement type is now significant when controlling for dose.

**Two-Factor ANOVA: Interaction Model (dose \* supp)**

```{r}
# Perform Two-Factor ANOVA (with interaction)
anova_interaction <- aov(len ~ dose * supp, data = tooth_data)
anova_interaction_summary <- summary(anova_interaction)

# Print formatted output
cat("\n----------------------------------------------------------\n")
cat("   TWO-FACTOR ANOVA: MODEL WITH INTERACTION (DOSE * SUPP)   \n")
cat("------------------------------------------------------------\n")
print(anova_interaction_summary)
```

**Interpretation: Dose and supplement type are both highly significant (p \< 0.001)**. **The interaction effect (dose:supp) is also significant (p = 0.02186)**, meaning the effect of dose depends on the supplement type.

### **Summary of ANOVA Results**

| Model | Significant Factors | Interpretation |
|------------------|------------------------------|------------------------|
| **One-Factor ANOVA (Dose)** | ✅ Dose (`p < 0.001`) | Higher doses significantly increase tooth growth. |
| **One-Factor ANOVA (Supplement)** | ❌ Supplement (`p = 0.0604`) | Supplement type **alone** is not a strong predictor. |
| **Two-Factor ANOVA (Dose + Supplement)** | ✅ Dose (`p < 0.001`), ✅ Supplement (`p = 0.0004`) | Both factors influence tooth growth, but dose has a stronger effect. |
| **Two-Factor ANOVA with Interaction** | ✅ Dose (`p < 0.001`), ✅ Supplement (`p < 0.001`), ✅ Interaction (`p = 0.0218`) | The effect of dose **depends on supplement type**. |

The **interaction plot** helps to determine **how the effect of dose on tooth growth depends on supplement type (OJ vs. VC).**

```{r}

# Generate interaction plot
ggplot(tooth_data, aes(x = dose, y = len, color = supp, group = supp)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line") +
  labs(title = "Interaction Plot: Effect of Dose and Supplement on Tooth Growth",
       x = "Dose (mg)", y = "Mean Tooth Length", color = "Supplement Type") +
  theme_minimal()
```

-   X-Axis: Dose levels (0.5 mg, 1 mg, 2 mg).

-   Y-Axis: Mean tooth length for each dose.

-   Red Line → Orange Juice (OJ).

-   Blue Line → Vitamin C (VC).

-   Data Points: Each dot represents the mean tooth length for that dose and supplement type.

-   Lines Connecting Points: Indicate the trend in tooth growth across increasing doses.

```{r}
# Compute summary statistics (mean & standard error)
summary_data <- tooth_data %>%
  group_by(dose, supp) %>%
  summarise(
    mean_len = mean(len),
    se = sd(len) / sqrt(n())  # Standard Error
  )

# View summary data
#print(summary_data)

# Bar plot with error bars
ggplot(summary_data, aes(x = dose, y = mean_len, fill = supp)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.6) +
  geom_errorbar(aes(ymin = mean_len - se, ymax = mean_len + se),
                width = 0.2, position = position_dodge(0.6)) +
  labs(
    title = "Effect of Dose and Supplement Type on Tooth Growth",
    x = "Dose (mg)", 
    y = "Mean Tooth Length (mm)", 
    fill = "Supplement Type"
  ) +
  theme_minimal()
```

This **bar plot** represents **mean tooth length** for different **doses** of **OJ (red) and VC (blue)**. **Error bars** represent **standard errors**, showing the variability of measurements. This visualization complements the interaction plot by providing a **clearer view of mean values and their variability**.

**Key Observations from the Bar Plot**

1.	**OJ results in greater tooth growth at lower doses**:

•	**At 0.5 mg, OJ (\~13 mm) results in significantly higher tooth growth than VC (\~8 mm)**.

•	**At 1 mg, OJ (\~22 mm) is still more effective than VC (\~16 mm)**.

2.	**At 2 mg, the difference disappears**:

•	**Both supplements result in similar tooth growth (\~26 mm)**, confirming that **dose dominates the effect at higher levels**.

3.	**The error bars show some variation**:

•	The standard errors indicate **how much individual measurements fluctuate** within each dose group.

•	**Higher doses show slightly more variability**, but the trend remains strong.

### **Post-hoc Comparisons (Tukey's HSD)**

Tukey’s **Honest Significant Difference (HSD) test** is a **post-hoc test** used after ANOVA to identify **which specific groups** differ significantly from each other. It controls for multiple comparisons and ensures **family-wise confidence** at **95%**.

```{r}
# Tukey's Honest Significant Difference Test
tukey_result <- TukeyHSD(anova_interaction)

# Print results
cat("\n------------------------------------------\n")
cat("     TUKEY MULTIPLE COMPARISONS OF MEANS    \n")
cat("--------------------------------------------\n")
print(tukey_result)
```

**Tukey’s HSD for Dose (dose factor)**

| Comparison | Difference (`diff`) | Lower Bound (`lwr`) | Upper Bound (`upr`) | Adjusted p-value (`p adj`) | Significance |
|------------|------------|------------|------------|------------|------------|
| **1 - 0.5 mg** | 9.13 | 6.36 | 11.90 | **0.000** | **Significant** |
| **2 - 0.5 mg** | 15.50 | 12.73 | 18.26 | **0.000** | **Significant** |
| **2 - 1 mg** | 6.37 | 3.60 | 9.13 | **0.000006** | **Significant** |

All dose comparisons are statistically significant (p \< 0.05). Increasing dose significantly increases tooth growth.

-   1 mg leads to a 9.13 mm increase over 0.5 mg.

-   2 mg leads to a 15.50 mm increase over 0.5 mg.

-   2 mg leads to a 6.37 mm increase over 1 mg.

Conclusion: Higher doses always result in significantly more tooth growth.

**Tukey’s HSD for Supplement Type (supp factor)**

| Comparison | Difference (`diff`) | Lower Bound (`lwr`) | Upper Bound (`upr`) | Adjusted p-value (`p adj`) | Significance |
|------------|------------|------------|------------|------------|------------|
| **VC - OJ** | -3.7 | -5.58 | -1.82 | **0.0002312** | **Significant** |

OJ results in significantly greater tooth growth than VC (p \< 0.001). The mean difference is 3.7 mm in favor of OJ.

Conclusion: OJ is more effective than VC at promoting tooth growth.

**Tukey’s HSD for Dose & Supplement Interaction (dose:supp)**

| Comparison | Difference (`diff`) | Lower Bound (`lwr`) | Upper Bound (`upr`) | Adjusted p-value (`p adj`) | Significance |
|------------|------------|------------|------------|------------|------------|
| **1:OJ - 0.5:OJ** | 9.47 | 4.67 | 14.27 | **0.0000046** | **Significant** |
| **2:OJ - 0.5:OJ** | 12.83 | 8.03 | 17.63 | **0.0000000** | **Significant** |
| **0.5:VC - 0.5:OJ** | -5.25 | -10.05 | -0.45 | **0.0242521** | **Significant** |
| **2:VC - 0.5:OJ** | 12.91 | 8.11 | 17.71 | **0.0000000** | **Significant** |
| **0.5:VC - 1:OJ** | -14.72 | -19.52 | -9.92 | **0.0000000** | **Significant** |
| **1:VC - 1:OJ** | -5.93 | -10.73 | -1.13 | **0.0073930** | **Significant** |
| **0.5:VC - 2:OJ** | -18.08 | -22.88 | -13.28 | **0.0000000** | **Significant** |
| **1:VC - 2:OJ** | -9.29 | -14.09 | -4.49 | **0.0000069** | **Significant** |
| **2:VC - 2:OJ** | 0.08 | -4.72 | 4.88 | **1.0000000** | **Not Significant** |

All comparisons involving dose show a significant difference (p \< 0.05), except for 2:VC - 2:OJ (p = 1.000), meaning that: OJ is more effective than VC at lower doses and at 2 mg, OJ and VC are equally effective.

Key Comparisons:

OJ at 1 mg is significantly better than OJ at 0.5 mg (p = 0.0000046).

OJ at 2 mg is significantly better than OJ at 0.5 mg (p = 0.0000000).

OJ at 2 mg is significantly better than VC at 0.5 mg (p = 0.0000000).

VC at 2 mg vs. OJ at 2 mg is NOT significantly different (p = 1.000).

Conclusion: At lower doses (0.5 mg and 1 mg), OJ is more effective than VC and at the highest dose (2 mg), OJ and VC are equally effective.

```{r}
# Convert Tukey HSD results into a dataframe for plotting
tukey_dose <- as.data.frame(tukey_result$dose)
tukey_supp <- as.data.frame(tukey_result$supp)
tukey_interaction <- as.data.frame(tukey_result$`dose:supp`)  # Ensure proper access

# Add comparison names
tukey_dose$Comparison <- rownames(tukey_dose)
tukey_supp$Comparison <- rownames(tukey_supp)
tukey_interaction$Comparison <- rownames(tukey_interaction)

# Rename columns for clarity
colnames(tukey_dose) <- c("Mean_Difference", "Lower_CI", "Upper_CI", "P_Value", "Comparison")
colnames(tukey_supp) <- c("Mean_Difference", "Lower_CI", "Upper_CI", "P_Value", "Comparison")
colnames(tukey_interaction) <- c("Mean_Difference", "Lower_CI", "Upper_CI", "P_Value", "Comparison")

# Combine all Tukey results
tukey_df <- rbind(tukey_dose, tukey_supp, tukey_interaction)

# View the final Tukey test results dataframe
print(tukey_df)

# Plot Tukey's HSD results
ggplot(tukey_df, aes(x = Mean_Difference, y = reorder(Comparison, Mean_Difference), fill = Mean_Difference > 0)) +
  geom_bar(stat = "identity", width = 0.6) +  # Bar plot
  geom_errorbar(aes(xmin = Lower_CI, xmax = Upper_CI), width = 0.2) +  # Confidence intervals
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +  # Reference line at zero
  labs(
    title = "Tukey HSD Test: Mean Differences with Confidence Intervals",
    x = "Mean Difference in Tooth Growth",
    y = "Comparison Groups"
  ) +
  scale_fill_manual(values = c("red", "blue"), guide = "none") +  # Colors for positive/negative differences
  theme_minimal()  # Minimalist theme
```

-   Bars represent mean differences in tooth growth for each comparison.

-   Error bars (horizontal lines) show 95% confidence intervals.

-   A vertical dashed line at zero helps identify non-significant differences.

-   Red bars indicate negative differences (e.g., VC leading to lower tooth growth than OJ).

-   Blue bars indicate positive differences (e.g., higher doses leading to more growth).

Higher doses result in significantly greater tooth growth.	OJ is significantly better than VC at lower doses. At 2 mg, both supplements are equally effective (bars overlapping zero). The confidence intervals confirm statistical significance.

Finally:

| Factor | Findings |
|-----------------------------------|-------------------------------------|
| Dose | Higher doses lead to significantly greater tooth growth (`p < 0.001`). |
| Supplement | OJ results in significantly higher tooth growth than VC (`p < 0.001`). |
| Interaction (Dose:Supp) | OJ is more effective than VC at lower doses (0.5 mg & 1 mg), but at 2 mg, both supplements perform equally. |

Practical Implications:

For maximizing tooth growth, increasing dose is the most effective approach.

For lower doses, OJ is the better supplement choice.

At high doses (2 mg), both OJ and VC are equally effective, so supplement choice does not matter.

